Early Stage Assessment

### 2018-11-27 papers/esa/esa.tex
Add comments to ESA

@@

We can the define the spherical harmonics, orthogonal on the unit sphere as \cite{DLMFSphericalandSpheroidalHarmonics, DLMFFerrers, DLMFAssociatedLegendre}:
\begin{align}
\Ylmfull &:= \clm \eimphi \Plm (\costheta) \\
&= \alphalm (1 - (\costheta)^2)^\frac{|m|}{2} \eimphi P^{(|m|,|m|)}_{l-|m|}(\costheta), \quad \text{where } 0 \le |m| \le l, \, l \in \No.
\end{align} 
Note that we can then express \(\Ylm\) in terms of \(x,y,z\) instead of \(\theta, \varphi\) by noting that \(\costheta = z\) and that \( \eimphi\) can be expressed in terms of \(x,y,z\) for any \(m\in\mathbb{Z}\). Indeed, they are polynomials in $x,y,z$ which we denote $\Ylm(x,y,z)$. They span all polynomials modulo the ideal generated by $x^2 + y^2 + z^2 - 1$.

\citep{dunkl2014orthogonal} provided a framework for gaining a recurrence relation for each \(\bigP_{l+1}\) in terms of the previous two sub-vectors \(\Pl\) and \(\bigP_{l-1}\), utilising the sub matrices that make up the block-tridiagnonal Jacobi operator matrices for multiplication by $x, y$ and $z$. Clenshaw's algorithm then gives us a way of evaluating a function given its expansion in the spherical harmonic basis (up to a finite order).

@@

For \(l \in \No\), \(m \in \Z\) s.t. \(0 \le |m| \le l\):
\begin{align}
x\,\Ylm(x,y,z) &= \Alm Y^{m+1}_{l+1}(x,y,z) +  \Blm Y^{m+1}_{l-1}(x,y,z) \nonumber \\
		     & \quad \quad \quad + \Dlm Y^{m-1}_{l+1}(x,y,z) + \Elm Y^{m-1}_{l-1}(x,y,z), \\
y\,\Ylm(x,y,z) &= - i \, \big[\Alm Y^{m+1}_{l+1}(x,y,z) +  \Blm Y^{m+1}_{l-1}(x,y,z) \big] \nonumber \\
		     &\quad \quad \quad + i \, \big[ \Dlm Y^{m-1}_{l+1}(x,y,z) + \Elm Y^{m-1}_{l-1}(x,y,z) \big], \\
z\,\Ylm(x,y,z) &= \Flm Y^{m}_{l+1}(x,y,z) + \Glm Y^{m}_{l-1}(x,y,z) ,
\end{align}
where \(\Alm, \Blm, \Dlm, \Elm, \Flm, \Glm \in \R\) have derivable and calculable formulae.

@@

\subsection{Evaluation of a scalar function on the sphere}
We can use the Clenshaw algorithm to evaluate a function at a given point \((x,y,z)\) on the unit sphere provided we know the coefficients of the function when expanded in the spherical harmonic basis, i.e. suppose \(f(x,y,z)\) is a function and we know the set \(\{\bold{f}_l\}\) s.t.
\begin{align}
f(x,y,z) \approx \sum^N_{l=0} \bold{f}_l^T \, \Pl (x,y,z), \quad \text{where } \Pl (x,y,z), \bold{f}_l \in \R^{2l+1} \text{ for each } l \in \{0,\dots,N\}.
\end{align}

@@

By noting that the tangent space is then spanned by $\{\gradYlm, \gradpYlm\}$, and that these gradients and perpendicular gradients are orthogonal polynomials themselves, we can use the same ideas to create a framework for functions in the tangent space as expansions in this OP basis.

@@

Then using the operators on the coefficient vectors, the timestepping method is: 
\begin{align}
\mathbf{u}^c_{n+1} &= \mathbf{u}^c_{n} + \Delta t \: (G \mathbf{h}^c_{n+1} - F K \mathbf{u}^c_{n+1}) \\
\mathbf{h}^c_{n+1} &= \mathbf{h}^c_n - \Delta t \: \mathcal{H} D \mathbf{u}^c_{n+1}
\end{align}
$\iff$
\begin{align}
\mathbf{u}^c_{n+1} &= (I + \Delta t F K + \Delta t^2 \mathcal{H} G D)^{-1} \: (\mathbf{u}^c_{n} + \Delta t \: G \mathbf{h}^c_n) \\
\mathbf{h}^c_{n+1} &= \mathbf{h}^c_n - \Delta t \: \mathcal{H} D \mathbf{u}^c_{n+1}
\end{align}

@@

As a stepping stone to working on spherical triangles, we will try to construct orthogonal polynomials on a circle arc, given by $x \in [h, 1]$, $y = (1-x^2)^\frac{1}{2}$. We will use the lanczos algorithm from ApproxFun \cite{ApproxFun} to obtain 1D OPs $\{T^h_n\}$ (orthogonal wrt $w(x) = (1-x^2)^{-\frac{1}{2}}$) and $\{U^h_n\}$ (orthogonal wrt $w(x) = (1-x^2)^{\frac{1}{2}}$) on the interval $[h,1]$, and create our 2D OPs according to:
\begin{align}
\bigP_0(x,y) := T^h_0(x) \equiv T_0,
\quad \quad 
\bigP_n(x,y) := \begin{bmatrix}
		T^h_n(x) \\
		y \: U^h_{n-1}(x)
	\end{bmatrix} \in \R^2 \quad \forall n \in \N,
\quad \quad 
\bigP := \begin{bmatrix}
		\bigP_0 \\
		\hline
		\bigP_1 \\
		\hline
		\bigP_2 \\
		\vdots \\
	\end{bmatrix}.
\end{align}

@@

By using a similar process to \cite{dunkl2014orthogonal} we can construct 2D orthogonal polynomials on the half disk from 1D orthogonal polynomials on the intervals \([0,1]\) and \([-1,1]\). We denote the half disk by \(\Omega\) where
\begin{align}
\Omega := \{(x,y) \in \R^2 \quad | \quad 0 < x < 1, \: -(1-x^2)^{\half} < y < (1-x^2)^{\half}\}.
\end{align}

@@

Fix \(a,b \in \R\). Then for any function \(f : \Omega \to \R\) we can approximate (or exactly express) \(f\) by
\begin{align}
f(x,y) + \sum_{n=0}^N \bigP_n^{(a,b)}(x,y)^T \: \mathbf{f}_n
\end{align}
where
\begin{align}
\bigP(x,y) := \begin{bmatrix}
		\bigP_0(x,y) \\
		\hline
		\bigP_1(x,y) \\
		\hline
		\bigP_2(x,y) \\
		\hline
		\vdots \\
		\hline
		\bigP_N(x,y)
	\end{bmatrix},
\quad
\bigP_n(x,y) := \begin{bmatrix}
		P_{n,0}(x,y) \\
		\vdots \\
		P_{n,n}(x,y)
	\end{bmatrix} \in \R^{n+1} \quad \forall n = 0,1,2,\dots,N,
\end{align}
and where
\begin{align}
\mathbf{f} &:= \begin{bmatrix}
		\mathbf{f}_0 \\
		\hline
		\mathbf{f}_1 \\
		\hline
		\mathbf{f}_2\\
		\hline
		\vdots \\
		\hline
		\mathbf{f}_N
	\end{bmatrix},
\quad
\mathbf{f}_n := \begin{bmatrix}
		f_{n,0} \\
		\vdots \\
		f_{n,n}
	\end{bmatrix} \in \R^{n+1} \quad \forall n = 0,1,2,\dots,N, \\
f_{n,k} &:= \frac{< f, \: \Pnk^{(a,b)} >_{W^{(a,b)}}}{|| \Pnk^{(a,b)} ||_{W^{(a,b)}}}
\end{align}

@@

We can obtain the matrix operator for \(\Delta\), that will take us from coefficients for expansion in the weighted space
\[
\Wii(x,y) \: \Pii(x,y)
\]
to coefficients in the non-weighted space
\[
\Pii(x,y).
\]
This operator will have Banded-Block-Banded structure. We first note that, for general \(a,b\) and general function \(f(x,y) = \mathbf{f}^T \: \bigPab(x,y)\), there exist Banded-Block-Banded matrix operators \(D_x^{(a,b)}, \: D_y^{(a,b)}, \: W_x^{(a,b)}, \: W_y^{(a,b)}\) such that:
\begin{align}
\dx[f(x,y)] &= \bigP^{(a+1,b+1)}(x,y)^T \: D_x^{(a,b)} \: \mathbf{f}, \\
\dy[f(x,y)] &= \bigP^{(a,b+1)}(x,y)^T \: D_y^{(a,b)} \: \mathbf{f}, \\
\dx[\Wab(x,y) \: f(x,y)] &= W^{(a-1,b-1)}(x,y) \: \bigP^{(a-1,b-1)}(x,y)^T \: W_x^{(a,b)} \: \mathbf{f}, \\
\dy[\Wab(x,y) \: f(x,y)] &= W^{(a,b-1)}(x,y) \: \bigP^{(a,b-1)}(x,y)^T \: W_y^{(a,b)} \: \mathbf{f},
\end{align}
Further, we note that there exist Banded-Block-Banded matrix operators \(T^{(1,0)\to(0,0)}\) and \(T^{(0,1)\to(1,1)}\) such that:
\begin{align}
x \bigP^{(1,0)}(x,y) &= \Big(T^{(1,0)\to(0,0)} \Big)^T \: \bigP^{(0,0)}(x,y) \\
\bigP^{(0,1)}(x,y) &= \Big(T^{(0,1)\to(1,1)} \Big)^T \: \bigP^{(1,1)}(x,y)
\end{align}
Then, the operator \(\Delta\) is given by
\begin{align}
    \Delta := D_x^{(0,0)} \: W_x^{(1,1)} + T^{(0,1)\to(1,1)} \: D_y^{(0,0)} \: T^{(1,0)\to(0,0)} \: W_y^{(1,1)}
\end{align}

@@

The goal here is to eventually investigate sparse spectral methods on triangulations of the sphere as an alternative to the spherical harmonics approach currently in use at the European Centre for Medium-range Weather Forecasts (ECMWF) in their weather and climate model. The aim would be to preserve the predictive skill of the current approach whilst avoiding the parallel scalability bottleneck from the global spectral transform, which is expected to inhibit the future performance of the ECMWF model.

>>>>

\sotodoinline{Outline the "big picture" here. Say that the ultimate goal is to solve PDEs on spherical triangles, and what is listed below are steps on the way to that goal.}

@@

We can the define the spherical harmonics, orthogonal on the unit sphere as \cite{DLMFSphericalandSpheroidalHarmonics, DLMFFerrers, DLMFAssociatedLegendre}:
\begin{align}
\Ylmfull &:= \clm \eimphi \Plm (\costheta) \\
&= \alphalm (1 - (\costheta)^2)^\frac{|m|}{2} \eimphi P^{(|m|,|m|)}_{l-|m|}(\costheta), \quad \text{where } 0 \le |m| \le l, \, l \in \No.
\end{align} 
Note that we can then express \(\Ylm\) in terms of \(x,y,z\) instead of \(\theta, \varphi\) by noting that \(\costheta = z\) and that \( \eimphi\) can be expressed in terms of \(x,y,z\) for any \(m\in\mathbb{Z}\). Indeed, they are polynomials in $x,y,z$ which we denote $\Ylm(x,y,z)$. They span all polynomials modulo the ideal generated by $x^2 + y^2 + z^2 - 1$. \sotodo{Mention something like: this is equivalent to standard techniques in the literature on spherical harmonics, but the goal here is to put it in a language that will translate to spherical triangles}

\citep{dunkl2014orthogonal} provided a framework for gaining a recurrence relation for each \(\bigP_{l+1}\) in terms of the previous two sub-vectors \(\Pl\) and \(\bigP_{l-1}\), utilising the sub matrices that make up the block-tridiagnonal Jacobi operator matrices for multiplication by $x, y$ and $z$. Clenshaw's algorithm then gives us a way of evaluating a function given its expansion in the spherical harmonic basis (up to a finite order).  \sotodo{Mention that "this provides an optimal complexity method for evaluating spherical harmonic expansions. This is an alternative to using the explicit expressions that may have benefits, though further investigation is required."}

@@

For \(l \in \No\), \(m \in \Z\) s.t. \(0 \le |m| \le l\):
\begin{align}
x\,\Ylm(x,y,z) &= \Alm Y^{m+1}_{l+1}(x,y,z) +  \Blm Y^{m+1}_{l-1}(x,y,z) \nonumber \\
		     & \quad \quad \quad + \Dlm Y^{m-1}_{l+1}(x,y,z) + \Elm Y^{m-1}_{l-1}(x,y,z), \\
y\,\Ylm(x,y,z) &= - i \, \big[\Alm Y^{m+1}_{l+1}(x,y,z) +  \Blm Y^{m+1}_{l-1}(x,y,z) \big] \nonumber \\
		     &\quad \quad \quad + i \, \big[ \Dlm Y^{m-1}_{l+1}(x,y,z) + \Elm Y^{m-1}_{l-1}(x,y,z) \big], \\
z\,\Ylm(x,y,z) &= \Flm Y^{m}_{l+1}(x,y,z) + \Glm Y^{m}_{l-1}(x,y,z) ,
\end{align}
where \(\Alm, \Blm, \Dlm, \Elm, \Flm, \Glm \in \R\) have derivable and calculable formulae.\sotodo{Mention that the formulae are explicit}

@@

\subsection{Evaluation of a scalar function on the sphere}
We can use the Clenshaw algorithm \sotodo{I would say "a multivariate analogue of Clenshaw's algorithm}  to evaluate a function at a given point \((x,y,z)\) on the unit sphere provided we know the coefficients of the function when expanded in the spherical harmonic basis, i.e. suppose \(f(x,y,z)\) is a function and we know the set \(\{\bold{f}_l\}\) s.t.
\begin{align}
f(x,y,z) \approx \sum^N_{l=0} \bold{f}_l^T \, \Pl (x,y,z), \quad \text{where } \Pl (x,y,z), \bold{f}_l \in \R^{2l+1} \text{ for each } l \in \{0,\dots,N\}.
\end{align}

@@

By noting that the tangent space is then spanned by\sotodo{Connect these to "vector spherical harmonics"} $\{\gradYlm, \gradpYlm\}$, and that these gradients and perpendicular gradients are orthogonal polynomials themselves, we can use the same ideas to create a framework for functions in the tangent space as expansions in this OP basis.

@@

Then using the operators on the coefficient vectors, the timestepping method is: \sotodo{Explain what $F$, $K$ , and $G$ are and emphasise they can all be constructed}
\begin{align}
\mathbf{u}^c_{n+1} &= \mathbf{u}^c_{n} + \Delta t \: (G \mathbf{h}^c_{n+1} - F K \mathbf{u}^c_{n+1}) \\
\mathbf{h}^c_{n+1} &= \mathbf{h}^c_n - \Delta t \: \mathcal{H} D \mathbf{u}^c_{n+1}
\end{align}
$\iff$
\begin{align}
\mathbf{u}^c_{n+1} &= (I + \Delta t F K + \Delta t^2 \mathcal{H} G D)^{-1} \: (\mathbf{u}^c_{n} + \Delta t \: G \mathbf{h}^c_n) \\
\mathbf{h}^c_{n+1} &= \mathbf{h}^c_n - \Delta t \: \mathcal{H} D \mathbf{u}^c_{n+1}
\end{align}

@@

As a stepping stone to working on spherical triangles, we will try to construct orthogonal polynomials on a circle arc, given by $x \in [h, 1]$, $y = (1-x^2)^\frac{1}{2}$. We will use the lanczos algorithm from\sotodo{"as implemented", not "from"} ApproxFun \cite{ApproxFun} to obtain 1D OPs $\{T^h_n\}$ (orthogonal wrt $w(x) = (1-x^2)^{-\frac{1}{2}}$) and $\{U^h_n\}$ (orthogonal wrt $w(x) = (1-x^2)^{\frac{1}{2}}$) on the interval $[h,1]$, and create our 2D OPs according to:\sotodo{Cite Huybrechs reference for connection in literature}
\begin{align}
\bigP_0(x,y) := T^h_0(x) \equiv T_0,
\quad \quad 
\bigP_n(x,y) := \begin{bmatrix}
		T^h_n(x) \\
		y \: U^h_{n-1}(x)
	\end{bmatrix} \in \R^2 \quad \forall n \in \N,
\quad \quad 
\bigP := \begin{bmatrix}
		\bigP_0 \\
		\hline
		\bigP_1 \\
		\hline
		\bigP_2 \\
		\vdots \\
	\end{bmatrix}.
\end{align}

@@

By using a similar process to\sotodo{Try to always include page numbers in citations to books, as otherwise you'll forget where it is} \cite{dunkl2014orthogonal} we can construct 2D orthogonal polynomials on the half disk from 1D orthogonal polynomials on the intervals \([0,1]\) and \([-1,1]\). We denote the half disk by \(\Omega\) where
\begin{align}
\Omega := \{(x,y) \in \R^2 \quad | \quad 0 < x < 1, \: -(1-x^2)^{\half} < y < (1-x^2)^{\half}\}.
\end{align}

@@

Fix \(a,b \in \R\). Then for any function \(f : \Omega \to \R\) we can approximate (or exactly express) \(f\) by\sotodo{The $+$ should be $\approx$}
\begin{align}
f(x,y) + \sum_{n=0}^N \bigP_n^{(a,b)}(x,y)^T \: \mathbf{f}_n
\end{align}
where
\begin{align}
\bigP(x,y) := \begin{bmatrix}
		\bigP_0(x,y) \\
		\hline
		\bigP_1(x,y) \\
		\hline
		\bigP_2(x,y) \\
		\hline
		\vdots \\
		\hline
		\bigP_N(x,y)
	\end{bmatrix},
\quad
\bigP_n(x,y) := \begin{bmatrix}
		P_{n,0}(x,y) \\
		\vdots \\
		P_{n,n}(x,y)
	\end{bmatrix} \in \R^{n+1} \quad \forall n = 0,1,2,\dots,N,
\end{align}
and where
\begin{align}
\mathbf{f} &:= \begin{bmatrix}
		\mathbf{f}_0 \\
		\hline
		\mathbf{f}_1 \\
		\hline
		\mathbf{f}_2\\
		\hline
		\vdots \\
		\hline
		\mathbf{f}_N
	\end{bmatrix},
\quad
\mathbf{f}_n := \begin{bmatrix}
		f_{n,0} \\
		\vdots \\
		f_{n,n}
	\end{bmatrix} \in \R^{n+1} \quad \forall n = 0,1,2,\dots,N, \\
f_{n,k} &:= \frac{\langle f, \: \Pnk^{(a,b)} \rangle_{W^{(a,b)}}}{|| \Pnk^{(a,b)} ||_{W^{(a,b)}}}
\end{align}

@@

We can obtain the matrix operator for \(\Delta\), that will take us from coefficients for expansion in the weighted space \sotodo{Mention this vanishes on the boundary so imposes zero Dirichlet}
\[
\Wii(x,y) \: \Pii(x,y)
\]
to coefficients in the non-weighted space\sotodo{The notation $\Pii$ is not defined I think}
\[
\Pii(x,y).
\]
This operator will have Banded-Block-Banded structure. We first note that, for general \(a,b\) and general function \(f(x,y) = \mathbf{f}^T \: \bigPab(x,y)\), there exist Banded-Block-Banded matrix operators \(D_x^{(a,b)}, \: D_y^{(a,b)}, \: W_x^{(a,b)}, \: W_y^{(a,b)}\) such that: \sotodo{Mention this is analoguous to Jacobi polynomials, DLMF ...}
\begin{align}
\dx[f(x,y)] &= \bigP^{(a+1,b+1)}(x,y)^T \: D_x^{(a,b)} \: \mathbf{f}, \\
\dy[f(x,y)] &= \bigP^{(a,b+1)}(x,y)^T \: D_y^{(a,b)} \: \mathbf{f}, \\
\dx[\Wab(x,y) \: f(x,y)] &= W^{(a-1,b-1)}(x,y) \: \bigP^{(a-1,b-1)}(x,y)^T \: W_x^{(a,b)} \: \mathbf{f}, \\
\dy[\Wab(x,y) \: f(x,y)] &= W^{(a,b-1)}(x,y) \: \bigP^{(a,b-1)}(x,y)^T \: W_y^{(a,b)} \: \mathbf{f},
\end{align}
Further, we note that there exist Banded-Block-Banded matrix operators \(T^{(1,0)\to(0,0)}\) and \(T^{(0,1)\to(1,1)}\) such that:
\begin{align}
x \bigP^{(1,0)}(x,y) &= \Big(T^{(1,0)\to(0,0)} \Big)^T \: \bigP^{(0,0)}(x,y) \\
\bigP^{(0,1)}(x,y) &= \Big(T^{(0,1)\to(1,1)} \Big)^T \: \bigP^{(1,1)}(x,y)
\end{align}
Then, the operator \(\Delta\) is given by \sotodo{Mention it is also Banded-Block-Banded, hence sparse}
\begin{align}
    \Delta := D_x^{(0,0)} \: W_x^{(1,1)} + T^{(0,1)\to(1,1)} \: D_y^{(0,0)} \: T^{(1,0)\to(0,0)} \: W_y^{(1,1)}
\end{align}

@@

The goal here is to eventually investigate sparse spectral methods on triangulations of the sphere as an alternative to the spherical harmonics approach currently in use at the European Centre for Medium-range Weather Forecasts (ECMWF) in their weather and climate model. The aim would be to preserve the predictive skill of the current approach whilst avoiding the parallel scalability bottleneck from the global spectral transform, which is expected to inhibit the future performance of the ECMWF model. \sotodo{This last paragraph could be in the intro}

\sotodoinline{Maybe mention that isosceles triangles should follow by the same procedure as the half sphere, but general spherical triangles will require further ideas.}

\sotodoinline{Expand bibliography: cite spherical harmonics books, Huybrechs, ECMWF paper, etc.}
